Modal operators for a constructive account of well quasi-orders

### 2025-02-25 slides-paris2025.tex
Slides finished

    \textbf{Def.} A (Grothendieck) \hil{topos} is a category equivalent to the
    category of sheaves over a small site.

    \textbf{Examples for toposes.}
    \begin{itemize}
      \item $\mathrm{Set}$, the category of sets and maps.
      \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
      \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
      filter} of a \hil{forcing notion} (a quasi-order equipped with a
      coverage).
    \end{itemize}
    The following are \bad{not} toposes:
    \begin{itemize}
      \item The category of sets and partially defined maps.
      \item The category of abelian groups.
    \end{itemize}

    \begin{enumerate}
      \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
      \medskip
      \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
      \medskip
      \item For every inhabited set~$X$, \emph{proximally} there exists an
      enumeration~$\NN \twoheadrightarrow X$.
      \medskip

      \item For every ring, \emph{proximally} there exists a maximal ideal.
      \medskip

      \item \emph{Somewhere}, the law of excluded middle holds.
    \end{enumerate}
  \end{changemargin}
\end{frame}}

>>>>

    \textbf{Def.} A (Grothendieck) \hil{topos} is a category equivalent to the
    category of sheaves over a small site.
    \pause

    \only<2>{
      \textbf{Examples for toposes.}
      \begin{itemize}
        \item $\mathrm{Set}$, the category of sets and maps.
        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
        \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
        filter} of a \hil{forcing notion} (a quasi-order equipped with a
        coverage).
      \end{itemize}
      The following are \bad{not} toposes:
      \begin{itemize}
        \item The category of sets and partially defined maps.
        \item The category of abelian groups.
      \end{itemize}
    }

    \only<3->{
      \hil{Multiversal yoga:}
      \begin{enumerate}
        \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
        \smallskip
        \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
        \smallskip
        \item For every inhabited set~$X$, \emph{proximally} there exists an
        enumeration~$\NN \twoheadrightarrow X$.
        \smallskip

        \item For every ring, \emph{proximally} there exists a maximal ideal.
        \smallskip

        \item \emph{Somewhere}, the law of excluded middle holds.
      \end{enumerate}
    }
  \end{changemargin}
\end{frame}}

\begin{frame}{Toposophic constructive combinatorics}
  \justifying
  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
  is well$\ind$.

  \emph{Proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
  arbitrary topos. We need to show that~$\alpha$ is good, so find indices~$n < m$ such that
  \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]

  It suffices to prove that \emph{somewhere}, $\alpha$ is good, as goodness is
  a geometric implication (in fact even a geometric formula). Hence without loss of
  generality, we may suppose~\textsc{lem}.

  Thus there is an infinite increasing subsequence
  \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
  As~$Y$ is well\ind, the sequence~$(\gamma\,k_0, \gamma\,k_1, \ldots)$
  is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
  also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Toposophic constructive algebra}
  \justifying
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

  \emph{Classical proof.}
  \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
  The matrix~$M$ is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \pause

  \emph{Multiversal constructive proof.}
  We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$.
  The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \pause

  \emph{Unrolled constructive proof (special case).}
  Write~$M =
  \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  have~$u, v$ with
  \vspace*{-0.6em}
  \[
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
    \quad\text{and}\quad
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
  \]
  Hence
  $
    1 = (vy) (ux) = (uy) (vx) = 0
  $.
  \qed
  \pause

  \centering
  \emph{\href{https://iblech.github.io/constructive-maximal-ideals/}{Agda
  formalization available.}}
  \par
\end{frame}}
